---
title: "K-Means and K-Medoids Clustering on Simulated Data"
author: "Halil Bisgin"
output: html_document
---

## Introduction

In this analysis, we simulate data to apply both **K-Means** and **K-Medoids** clustering algorithms, compare their results, and visualize the clustering outcomes.

We also measure the time taken for each clustering method.

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
# Install libraries if not already installed
 install.packages("cluster")
 install.packages("factoextra")
 install.packages("ggplot2")
 install.packages("tictoc")

# Load necessary libraries
library(cluster)      # For K-Medoids (PAM)
library(factoextra)   # For visualization
library(ggplot2)      # For plotting
library(tictoc)       # For measuring execution time
```

```{r data prep}
# Set random seed for reproducibility
set.seed(42)

# Create 3 clusters with normal distribution
n <- 1500  # Total number of data points
cluster1 <- data.frame(x = rnorm(n/3, mean = 2, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
cluster2 <- data.frame(x = rnorm(n/3, mean = 6, sd = 0.5), y = rnorm(n/3, mean = 6, sd = 0.5))
cluster3 <- data.frame(x = rnorm(n/3, mean = 10, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))

# Combine clusters into one dataset
data <- rbind(cluster1, cluster2, cluster3)
colnames(data) <- c("feature1", "feature2")

# Normalize the data
scaled_data <- scale(data)

```
# Apply K-Means Clustering

```{r k-means}
tic("K-Means")  # Start timer
kmeans_result <- kmeans(scaled_data, centers = 3, nstart = 10)
toc()  # Stop timer
```
# K-means when n = 300 elapsed 0.03 seconds
# K-Means when n = 500 elapsed 0.02 seconds
# K-means when n = 700 elapsed 0.02 seconds
# K-means when n = 900 elapsed 0.03 seconds
# K-means when n = 1100 elapsed 0.06 seconds
# K-means when n = 1300 elapsed 0.04 seconds
# K-means when n = 1500 elapsed 0.04 seconds

# Apply K-Medoids Clustering
```{r k-medoids}
tic("K-Medoids")  # Start timer
kmedoids_result <- pam(scaled_data, 3)
toc()  # Stop timer

```
# K-medoids when n = 300 elapsed 0.03 seconds
# K-medoids when n = 500 elapsed 0.05 seconds
# K-medoids when n = 700 elapsed 0.07 seconds
# K-medoids when n = 900 elapsed 0.08 seconds
# K-medoids when n = 1100 elapsed 0.12 seconds
# K-medoids when n = 1300 elapsed 0.12 seconds
# K-medoids when n = 1500 elapsed 0.23 seconds

# Prep data for visualization
```{r get memberships }
data$KMeans_Cluster <- as.factor(kmeans_result$cluster)
data$KMedoids_Cluster <- as.factor(kmedoids_result$clustering)
```

# Plotting k-means clusters
```{r vis kmeans}
p1 <- ggplot(data, aes(x = feature1, y = feature2, color = KMeans_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Means Clustering") +
  theme_minimal()

# Print the K-Means plot
print(p1)
```

# Plotting k-medoids clusters
```{r kmedoids}
p2 <- ggplot(data, aes(x = feature1, y = feature2, color = KMedoids_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Medoids Clustering") +
  theme_minimal()

# Print the K-Medoids plot
print(p2)
```